The Courtroom Challenge—A Conjecture on the Unavoidable Cost of LLM Verification

### 2025-06-06 main.tex
Add files via upload

% ---------- title data ----------
\title{The Courtroom Challenge\\\large Can Any Open-World AI Thrive Without Paying the Validation Bill?\\[1em]\small\textit{This note poses a conjecture, not a leaderboard: automate validation, prove it's unnecessary, or price human judgment honestly.}}
\author{John Repsys\thanks{email: john.repsys@gmail.com}}
\date{\today}

@@

% =========================================================
\begin{document}
\maketitle
\begin{abstract}
  Large language models hallucinate; ad-hoc patches conceal the true cost of verification. We therefore pose the Courtroom Challenge:
  Conjecture 1. For any unbounded open-world text stream, a system that lacks both (i) a persistent precedent store and (ii) a friction-priced validation loop will accumulate unbounded prequential regret.
  We formalise the state machine, cost ledger, and impossibility triangle that make this conjecture falsifiable, and invite the community to break it. A 100-query toy stream illustrates how the ledger is logged\footnote{Baseline script will appear at \url{https://github.com/johnrepsys/courtroom-model-lean} with v2.}; no attempt is made here to optimise the validator. In addition, we present the Courtroom Model, a wrapper architecture that subjects every LLM claim to a perpetual cycle of precedent search, automated semantic checks, and selective escalation to an external judge—human or automated. Each new output is validated against a persistent vector memory, while conflicting evidence triggers the re-examination of prior “case law.”
\end{abstract}

@@

Cosine and L$_2$ are special cases, but a system designer is free to
plug in any monotone metric—e.g.\ Mahalanobis distance, dot-product in a
learned embedding space, or even a hybrid score that blends semantic and
numeric keys—so long as the inequality is flipped to match the
metric's notion of proximity. Section~\ref{sec:toy-storage-bound} analyses the cost/error trade-off
under the distance form (Lemma~\ref{lemma}), and Appendix~\ref{sec:knobs} lists this dial
among the falsifiability-critical degrees of freedom. Selecting which descriptor 
fields to embed is itself tunable; Section~\ref{sec:open-questions} (Q\ref{open-question-schema}) sketches how the Courtroom loop 
could learn or prune schema columns over time.

@@

\textbf{Classical identification bound.}
Choosing the 0–1 loss $L(M,X)=\mathbf 1\{\hat M(X)\neq M\}$
and threshold $\Delta=\tfrac12$
in Proposition 3 of \citet{chen2024interfano} yields
\[
  P_{\mathrm{id}}
  \;\ge\;
  \tfrac12\Bigl(1-\tfrac{I+\log 2}{\log M}\Bigr).
\]

\medskip
\noindent
To obtain a bound on label error we apply the same proposition to
the indicator
\[
  L^\star \;=\; \mathbf 1_{\{\Err>\Delta_{\mathrm{Err}}\}},
  \qquad
  \Delta_{\mathrm{Err}} := \tfrac12\,\delta\,\kappabar\,T.
\]

\item
  Re-apply Proposition 3 to
  $L^\star = \mathbf 1\{\Err>\Delta_{\mathrm{Err}}\}$
  with threshold $\Delta=\tfrac12$.
  Because $\log \!\bigl(1/\rho_{1/2}\bigr)\ge\log M$ for the uniform
  prior, the proposition yields the lower bound
  \[
    \Prb\!\bigl[\Err>\Delta_{\mathrm{Err}}\bigr]
      \;\ge\;
      \frac12\Bigl(
        1-\frac{I+\log 2}{\log M}
      \Bigr).
  \]

@@

\subsection{Toy Storage Bound}
\label{sec:toy-storage-bound}

@@

\begin{lemma}[Cost–error lower bound]\label{lem:cost-error}
  For any query stream of min-entropy $H>0$, any validator whose
  \emph{amortised escalation cost} is $o(1)$ must incur
  \emph{per-query error} $\Omega(1)$ against an adaptive adversary.
\end{lemma}

\begin{proof}[Proof sketch]
  The adversary at round $t$ submits the first query not present in
  the current precedent.  Since the validator escalates $o(T)$ times
  over $T$ rounds, at least $T - o(T)$ queries are unseen, forcing a
  coin-flip and hence expected error $\ge \tfrac12$. \footnote{A fully mechanised Lean proof of Lemma 2 will appear at \url{https://github.com/johnrepsys/courtroom-model-lean}.}
\end{proof}

>>>>

% ---------- title data ----------
\title{The Courtroom Challenge—A Conjecture on the Unavoidable Cost of LLM Verification}
\author{John Repsys\thanks{email: john.repsys@gmail.com}}
\date{\today}

@@

% =========================================================
\begin{document}
\maketitle
\begin{abstract}
  Large language models hallucinate; ad-hoc patches conceal the true cost of verification. 
  We therefore pose a falsifiable ultimatum: either automate validation, prove it unnecessary, or pay the explicit human-in-the-loop cost.  
  Conjecture 1 (the Courtroom Challenge). 
  For any unbounded open-world text stream, a system that lacks both (i) a persistent precedent store and (ii) a friction-priced validation loop will accumulate unbounded prequential regret.
  We formalise the state machine, cost ledger, and impossibility triangle that make this conjecture testable, and invite the community to break it.
  A 100-query toy stream illustrates how the ledger is logged\footnote{Baseline script will appear at \url{https://github.com/johnrepsys/courtroom-model-lean} with v2.}; no attempt is made here to optimise the validator.
  In addition, we present the Courtroom Model, a wrapper architecture that subjects every LLM claim to a perpetual cycle of precedent search, automated semantic checks, and selective escalation to an external judge—human or automated.
  Each new output is validated against a persistent vector memory, while conflicting evidence triggers the re-examination of prior “case law.”
  The challenge stands: thrive in an open world without paying the validation bill—or show why the bill is inescapable.
\end{abstract}

@@

Cosine and L$_2$ are special cases, but a system designer is free to
plug in any monotone metric—e.g.\ Mahalanobis distance, dot-product in a
learned embedding space, or even a hybrid score that blends semantic and
numeric keys—so long as the inequality is flipped to match the
metric's notion of proximity. Section~\ref{sec:sparse-memory-bound} analyses the cost/error trade-off
under the distance form (Lemma~\ref{lemma}), and Appendix~\ref{sec:knobs} lists this dial
among the falsifiability-critical degrees of freedom. Selecting which descriptor 
fields to embed is itself tunable; Section~\ref{sec:open-questions} (Q\ref{open-question-schema}) sketches how the Courtroom loop 
could learn or prune schema columns over time.

@@

\textbf{Classical identification bound.}
Choosing the 0–1 loss $L(M,X)=\mathbf 1_{\{\hat M(X)\neq M\}}$
and threshold $\Delta=\tfrac12$
in Proposition 3 of \citet{chen2024interfano} yields
\[
  P_{\mathrm{id}}
  \;\ge\;
  \tfrac12\Bigl(1-\tfrac{I+\log 2}{\log M}\Bigr).
\]

\medskip
\noindent
To obtain a bound on label error we apply the same proposition with 
threshold $\Delta=\tfrac12$ to the indicator
\[
  L^\star \;=\; \mathbf 1_{\{\Err>\Delta_{\mathrm{Err}}\}},
  \qquad
  \Delta_{\mathrm{Err}} := \tfrac12\,\delta\,\kappabar\,T.
\]

\item
  Because $\log \!\bigl(1/\rho_{1/2}\bigr)\ge\log M$ for the uniform
  prior, the proposition yields the lower bound
  \[
    \Prb\!\bigl[\Err>\Delta_{\mathrm{Err}}\bigr]
      \;\ge\;
      \frac12\Bigl(
        1-\frac{I+\log 2}{\log M}
      \Bigr).
  \]

@@

\subsection{Sparse-memory bound}
\label{sec:sparse-memory-bound}

The following lemma gives a back-of-the-envelope bound on memory growth and justifies the choice of a retention threshold $\tau$.

@@
